Induced universal graphs for families of small graphs

### 2021-10-15 paper/paper.tex
Writing about strategies

\usepackage[letterpaper, total={6in, 8in}]{geometry}
\usepackage[x11names, svgnames, rgb]{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{amsthm}
\usetikzlibrary{decorations, decorations.pathreplacing, decorations.pathmorphing,
    calc, backgrounds, positioning, tikzmark, patterns, matrix, fit, graphs,
    snakes, arrows, shapes, calligraphy}
%\usetikzlibrary{snakes,arrows,shapes}
\usepackage{amsmath}
\usepackage{subfigure}
\usepackage{cleveref}

@@

\newcommand{\calF}{\ensuremath{\mathcal{F}}}

@@

Given $k$ and $n$,
I used a simple brute-force approach to find the set of
$n$-vertex graphs that are induced universal for the family $\calF(k)$ of all $k$-vertex
graphs.  Each possible $n$-vertex graph $G$ from Brendan McKay's lists of
graphs\footnote{\url{https://users.cecs.anu.edu.au/~bdm/data/graphs.html}} was
tested in turn.  I used a Python implementation of the McSplit
algorithm \cite{DBLP:conf/ijcai/McCreeshPT17} to test whether each $k$-vertex
graph is an induced subgraph of $G$.  Although the McSplit algorithm was designed for
the more general maximum common induced subgraph problem, it performs well on
the subgraph isomorphism problem if both input graphs are small.  (Although
I have described the application of this method to families of all graphs of a given order,
the method could easily be applied to any family $\calF$ of graphs, and indeed
Section~\ref{sec:trees} gives results from the application of this method
to families of trees.)

For those graphs $G$ that do not contain a copy of each element of $\calF(k)$,
it improves the speed of our algorithm if we can iterate over
$\calF(k)$ in an order that allows us to quickly reject $G$ by finding a graph
that is not an induced subgraph of $G$.  The approach I used was to sort
$\calF(k)$ in descending order of number of isomorphisms.  In particular, this
means that the clique $K_k$ and the independent set $I_k$ are the first two
graphs tested.

@@

>>>>

\usepackage[letterpaper, total={6in, 8in}]{geometry}
\usepackage[x11names, svgnames, rgb]{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{amsthm}
\usetikzlibrary{decorations, decorations.pathreplacing, decorations.pathmorphing,
    calc, backgrounds, positioning, tikzmark, patterns, matrix, fit, graphs,
    snakes, arrows, shapes, calligraphy}
%\usetikzlibrary{snakes,arrows,shapes}
\usepackage{amsmath}
\usepackage{subfigure}
\usepackage{cleveref}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{algpseudocode}

@@

\crefname{algorithm}{Algorithm}{Algorithms}
\Crefname{algorithm}{Algorithm}{Algorithms}
\crefname{algocf}{Algorithm}{Algorithms}
\Crefname{algocf}{Algorithm}{Algorithms}
\crefname{figure}{Figure}{Figures}
\Crefname{figure}{Figure}{Figures}
\crefname{table}{Table}{Tables}
\Crefname{table}{Table}{Tables}
\crefname{section}{Section}{Sections}
\Crefname{section}{Section}{Sections}

\newcommand{\calF}{\ensuremath{\mathcal{F}}}
\newcommand{\calG}{\ensuremath{\mathcal{G}}}
\newcommand{\AlgVar}[1]{\mathit{#1}}

\newcommand{\lineref}[1]{line~\ref{#1}}
\newcommand{\codelineref}[1]{line~\ref{#1}}
\newcommand{\linerangeref}[2]{lines~\ref{#1} to~\ref{#2}}
\newcommand{\Lineref}[1]{Line~\ref{#1}}
\newcommand{\Linerangeref}[2]{Lines~\ref{#1} to~\ref{#2}}

% https://tex.stackexchange.com/a/184829
\algnewcommand{\LeftComment}[1]{\(\triangleright\) #1}

@@

Given $\calF$ and $n$,
we use the brute-force approach described by Preen \cite{preen_math_se} and shown
in \Cref{alg:brute-force} to find the set of
$n$-vertex graphs that are induced universal for the family $\calF$.
The function $\FuncSty{AllInducedUniversalGraphs}()$ is the entry point.
\Lineref{TryEachG} tests in turn each candidate graph $G$ from the family
of $n$-vertex graphs, and adds to the collection $\calG$ those that are induced universal
for $\calF$.  The function $\FuncSty{IsInducedUniversal}()$ tests
whether a graph $G$ is induced universal for $\calF$ by checking that
$H$ is isomorphic to a subgraph of $G$ for each $G \in \calF$.

For the collection $\calF(n)$ of candidate graphs on \lineref{TryEachG}, our
implementation uses Brendan McKay's lists of
graphs\footnote{\url{https://users.cecs.anu.edu.au/~bdm/data/graphs.html}}.
These are read in \texttt{graph6}
format\footnote{\url{https://users.cecs.anu.edu.au/~bdm/data/formats.html}}
from a text file as the program proceeds, and therefore only one candidate
graph at a time needs to be stored in memory.

For the call $\FuncSty{InducedSubIso}()$ to an algorithm for the induced
subgraph isomorphism decision problem on line \lineref{CallSubIso},
our program uses a Python implementation of the McSplit
algorithm \cite{DBLP:conf/ijcai/McCreeshPT17}.
McSplit was designed for
the more general maximum common induced subgraph problem,
but it can be used for induced subgraph isomorphism with a simple
modification: we backtrack when the calculated upper bound
on the order of a common subgraph is less the order of graph $H$.
This method for solving
the subgraph isomorphism problem is very fast if both input graphs
are small, as the graphs considered in this paper are.
While McSplit is suitable for our purposes in this paper,
we do not claim that it is the the fastest induced subgraph isomorphism
solver for very small graphs; it would be useful future work to perform
an experimental comparison with other subgraph isomorphism solvers.
%%% (Although
%%% I have described the application of this method to families of all graphs of a given order,
%%% the method could easily be applied to any family $\calF$ of graphs, and indeed
%%% Section~\ref{sec:trees} gives results from the application of this method
%%% to families of trees.)

\begin{algorithm}[h!]
\DontPrintSemicolon
%%% \nl $\FuncSty{Search}(\AlgVar{future},M)$ \;
%%% \nl \Begin{
%%% %\nl \lIf {$\AlgVar{future} = \emptyset$ \bf{and} $|M| > |\AlgVar{incumbent}|$}
%%% \nl \lIf {$|M| > |\AlgVar{incumbent}|$}{$\AlgVar{incumbent} \gets M$} \label{StoreIncumbent}
%%% %\nl \lIf {$\AlgVar{future} = \emptyset$}{return}
%%% \medskip
%%% \nl $\AlgVar{bound} \gets |M|  + \sum_{\langle \setG,\setH \rangle \in \AlgVar{future}} \min(|\setG|,|\setH|)$ \label{CalcBound} \;
%%% \nl \lIf {$\AlgVar{bound} \leq |\AlgVar{incumbent}|$}{\KwSty{return}} \label{PruneSearch}
%%% \medskip
%%% \nl $\langle \setG,\setH \rangle \gets \FuncSty{SelectLabelClass}(\AlgVar{future})$ \label{SelectClass} \;
%%% \nl $v \gets \FuncSty{SelectVertex}(\setG)$ \label{SelectVertex} \;
%%% \nl \For {$w \in \setH$ \label{WLoop}} {
%%% \nl    $\AlgVar{future'} \gets \emptyset$ \label{NewFuture} \;
%%% \nl    \For {$\langle \setG',\setH'\rangle \in future$ \label{InnerLoop}}{
%%% \nl        $\setG'' \gets \setG' \cap \N(\graphG, v) \setminus \{v\}$ \label{NewPWithEdge} \;
%%% \nl        $\setH'' \gets \setH' \cap \N(\graphH, w) \setminus \{w\}$ \;
%%% \nl        \If {$\setG'' \neq \emptyset$ \bf{and} $\setH'' \neq \emptyset$\label{IfNonEmpty}}{
%%% \nl            $\AlgVar{future'} \gets \AlgVar{future'} \cup \{\langle \setG'' , \setH'' \rangle\}$ \label{AddToFutureWithEdge}}
%%% \nl        $\setG'' \gets \setG' \cap \invN(\graphG, v) \setminus \{v\}$ \label{NewPWithoutEdge}  \;
%%% \nl        $\setH'' \gets \setH' \cap \invN(\graphH, w) \setminus \{w\}$ \;
%%% \nl        \If {$\setG'' \neq \emptyset$ \bf{and} $\setH'' \neq \emptyset$\label{IfNonEmpty2}}{
%%% \nl            $\AlgVar{future'} \gets \AlgVar{future'} \cup \{\langle \setG'' , \setH'' \rangle\}$} \label{InnerLoopEnd}
%%%        }
%%% \nl   $\FuncSty{Search}(\AlgVar{future'},M\cup \{(v,w)\})$ \label{ExpandWithV} \;
%%%   }
%%% \nl $\setG' \gets \setG \setminus \{v\}$ \label{RemoveV} \;
%%% \nl $\AlgVar{future} \gets \AlgVar{future} \setminus \{\langle \setG,\setH \rangle\}$\;
%%% \nl \lIf {$\setG' \neq \emptyset$} {$\AlgVar{future} \gets \AlgVar{future} \cup \{\langle \setG',\setH \rangle \}$}
%%% \nl $\FuncSty{Search}(\AlgVar{future},M)$ \label{ExpandWithoutV} \;
%%% }
%%% \;
\nl $\FuncSty{IsInducedUniversal}(\calF,G)$ \label{BruteForceFun} \;
\nl \KwData{A family of graphs $\calF$ and a graph $G$.}
\nl \KwResult{A logical value indicating whether $G$ is induced universal for $\calF$.}
\nl \Begin{
\nl    \For {$H \in \calF$ \label{HLoop}}{
\nl      \lIf {$\neg\FuncSty{InducedSubIso}(H, G)$\label{CallSubIso}}{$\KwSty{return}$~$\AlgVar{false}$}
       }
\nl   $\KwSty{return}$~$\AlgVar{true}$ \;
    }
\medskip
\nl $\FuncSty{AllInducedUniversalGraphs}(\calF,n)$ \label{BruteForceFun} \;
\nl \KwData{A family of graphs $\calF$ and a natural number $n$.}
\nl \KwResult{The set of all graphs of order $n$ that are induced universal for $\calF$.}
\nl \Begin{
\nl    $\calG \gets \emptyset$ \;
\nl    \For {$G \in \calF(n)$}{
\nl      \lIf {$\FuncSty{IsInducedUniversal}(\calF,G)$\label{TryEachG}}{$\calG \gets \calG \cup \{G\}$}
       }
\nl   $\KwSty{return}$~$\calG$ \;
    }
\caption{A brute-force algorithm for finding all order-$n$ induced universal subgraphs of a family $\calF$ of graphs.
	The entry point is $\FuncSty{AllInducedUniversalGraphs}(\calF,n)$.}
\label{alg:brute-force}
\end{algorithm}

@@

\section{Iteration order for $\calF$}\label{sec:iteration-order}

When deciding on \lineref{HLoop} of the algorithm whether a graph $G$ contains
every element of $\calF$ as an induced subgraph, are some iteration orders of
$\calF$ better than others?  For a graph $G$ that contains a copy of each
element of $\calF$, the iteration order is of no importance; we much check
every element of $\calF$ before returning $\AlgVar{true}$.  For a graph $G$ that
does \emph{not} contain a copy of each element of $\calF$, however, we would like
to iterate over $\calF$ in an order such that graphs that are likely to fail the
subgraph isomorphism test are checked early; that way, we can quickly return
$\AlgVar{false}$.

We begin by considering the case where $\calF = \calF(k)$ for some $k$; that is,
$\calF$ is the set of all graphs of order $k$.
As Diaconis and Chatterjee note, \cite{chatterjee2021isomorphisms}, the complete
graph $K_n$ is the `hardest' graph of order $n$ for a random graph to contain.
This led me
to consider two possible orderings in which $K_k$ and its complement $I_k$ come
first.  The first of these approaches is to the sort $\calF(k)$ such that
graphs with large automorphism groups appear first.  (Intuitively, if a graph
has few automorphisms then we can generate many different labelled graphs
by permuting the vertex labels.  This gives more `opportunities' for the
graph to be an induced subgraph of $G$.)
The second approach considered is to place graphs
with unusually high or low edge counts, as measured by the absolute value of
${2|E(G)| - {|V(G)| \choose 2}}$, first.

\Cref{fig:scatter} examines whether graphs that appear early in the orderings
of $\calF$
generated by each of these two strategies are contained in few graphs with
a given number of vertices, as we would hope to be the case.
For this experiment, we let $\calF = \calF(5)$, and we consider the problem
of finding induced universal graphs of order 8.
Each point on the plots
represents one of the 34 graphs of order 5 (fewer than 34 points are visible
due to exactly-overlapping points, as we would expect).  On the vertical
axis, we have the number of graphs in $\calF(8)$ containing $G$; we would like
graphs for which this value is small to appear early in our order.
On the horizontal axes we have each of our two measures: size of automorphism group
and `extremeness' of edge count.  The first of these measures has a strong correlation
with the number of graphs containing $G$ as an induced subgraph, suggesting that
it is a good strategy to use when sorting the list of graphs.

%    \subfigure[][TODO caption] {
      \includegraphics*{img/automorphisms-scatter.pdf}
%      \label{fig:scatter-a}
%    }
     \qquad
%    \subfigure[][TODO caption] {
      \includegraphics*{img/deg-scatter.pdf}
%      \label{fig:scatter-b}
%    }
    \caption{For each graph $G \in \calF(5)$, we plot on the vertical axis
    the number of graphs in $\calF(8)$ that contain $G$ as an induced subgraph.
    This value has a strong negative correlation with the size of the automorphism
    group of $G$ (left plot), but a weaker correlation with a measure of how
    extreme the edge count of $G$ is (right plot).
    All axis scales are logarithmic.}
\label{fig:scatter}
\end{figure}

\subsection{Testing the strategies: $\calF = \calF{5}$}

We now test whether each of our ordering strategies results, as hoped,
in a reduced number of calls to the subgraph isomorphism function
in \Cref{alg:brute-force}.  As our first test case, we consider 
$\FuncSty{AllInducedUniversalGraphs}(\calF(5),9)$ ---
finding all induced universal graphs of order 9 for the family of
all graphs of order 5.  We consider four strategies.
The first two of these are the automorphism-count and `degree extremeness'
measures described above.  The fourth strategy is a random order.
Finally, we used one further strategy, which we call ``almost random''.
Under this strategy, $I_5$ and $K_5$ are placed at the start of the list,
and the remaining 32 graphs are in random order.  This strategy was not
intended to be useful in practice, but to give insight into whether
the success of the first two strategies was purely due to having
$I_5$ and $K_5$ at the start of the list of graphs.
For each of these strategies, we ran the program 50 times.

The results are shown in \Cref{fig:second-experiment}.  Over 50 runs,
the `automorphisms` strategy required a mean of 304,758 calls to the subgraph
isomorphism solver, while the `degree' strategy required a very similar
number of calls:  a mean of 304,701.  Surprisingly, the `almost random'
strategy had the lowest average number of calls of all the strategies,
at 304,431.  The `random' strategy fared poorly, requiring 822,649
calls on average and a minimum of 451,287 calls in the best of the 50 runs.
The variance in number of calls was tiny for all but the random strategy,
with between 304,000 and 305,000 calls being required for each run.

We note, first, that `degree' strategy appears to be at least as effective
as the `automorphisms' strategy for this instance, despite the lower
correlation observed in \Cref{fig:scatter}.  Perhaps most surprising is
that the `almost random' strategy performed better than any of the other
strategies.

    \includegraphics*[width=0.7\textwidth]{img/second-experiment-plot}

    \caption{TODO}
\label{fig:second-experiment}
\end{figure}

\subsection{Testing the strategies: $\calF \subset \calF{5}$}

TODO abcdef

    \includegraphics*[width=0.7\textwidth]{img/second-experiment-plot-using-sample}

    \caption{TODO}
\label{fig:second-experiment-using-sample}
\end{figure}
